Implementing Equational Constraints\\ in a Functional Language

### 2011-07-11 papers/wlp2011/paper.tex
benchmark update

In general, to extract all values from a \code{Choice} structure,
we have to traverse it, compute all possible choices
but consider the choice identifiers to make consistent (left/right)
decisions.
The latter requirement can be implemented by storing
the decisions already made for some choices during the traversal.
For this purpose, we introduce the type
\begin{haskell}
  data Decision = NoDecision | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoDecision} represents the fact that the value of a choice
is yet undecided.
Furthermore, we assume operations to lookup the current decision
for a given identifier or change its decision (depending on the implementation
of \code{IDSupply}, KiCS2 supports various implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupDecision :: ID -> IO Decision
  setDecision    :: ID -> Decision -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following I/O operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
  printValsDFS v                = print v
   where
    follow ChooseLeft  = printValsDFS x1
    follow ChooseRight = printValsDFS x2
    follow NoDecision  = do newDecision ChooseLeft  x1
                            newDecision ChooseRight x2

    newDecision d x = do setDecision i d
                         printValsDFS x
                         setDecision i NoDecision
\end{haskell}
This operation ignores failures and prints values that are not a \code{Choice}.
For a \code{Choice}, it checks whether a decision for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoDecision}).
If a decision has been made for this choice, it follows this decision.
Otherwise, the left alternative is used and this decision is stored.
After printing
all the values w.r.t.\ this decision,
the decision is undone (like in backtracking)
and the right alternative is used and stored.

@@

  Fail         & _     =  Fail
  _            & Fail  =  Fail
  Choice i l r & c     =  Choice i (l & c) (r & c)
  _            & _     =  Fail

@@

In this section we evaluate our implementation by some benchmarks.
They were executed on a Linux machine running Debian 5.0.7 with
an Intel Core 2 Duo (3.0GHz) processor.
KiCS2 has been executed with
the Glasgow Haskell Compiler (GHC 7.0.4, option -O2).
For a comparison with other mature implementations of Curry,
we considered PAKCS \cite{Hanus10PAKCS}
(Version 1.9.2, based on a SICStus-Prolog 4.1.2)
and MCC \cite{Lux99FLOPS} (Version 0.9.10). The timings were performed with the
time command measuring the execution time (in seconds) of a compiled
executable for each benchmark as a mean of three runs.
The programs used for the benchmarks,
partially taken from \cite{AntoyHanus05LOPSTR},
are \code{last} (compute the last element of a list),\footnote{%
\ccode{inc x n} is a naive addition that \code{n} times
increases its argument \code{x} by 1.}
\code{simplify} (simplify a symbolic arithmetic expression),
\code{varInExp} (non-deterministically return a variable occuring
                  in a symbolic arithmetic expression),
\code{half} (compute the half of a peano number using logic variables),
and
\code{palindrome} (check whether a list is a palindrome).

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & \code{==~}& \code{=:=} & \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..100000])}   & 118.33    & 0.52       & 0.20        \\
% \code{grep}                             & 0.94      & 0.13       &             \\
\code{simplify}                         & 9.18      & 10.33      & 7.09        \\
\code{varInExp}                         & 0.50      & 86.27      & 0.14        \\
\code{fromPeano (half (toPeano 10000))} & 26.49     & 6.26       & 11.58       \\
\code{palindrome}                       & 31.12     & 13.45      & 17.55       \\
% \code{horseman}                         & 3.88      & 4.93       &             \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different equation solving methods}
 \label{fig:search-strategies}
\end{figure}

In Section~\ref{sec:Unification} we mentioned that equational
constraints could also be solved by generators without variable bindings,
but this technique might increase the search space due to the
possibly superfluous generation of all values.
To show the beneficial effects of our implementation
of equational constraints with variable bindings,
we compare in Fig.~\ref{fig:search-strategies}
the results of using equational constraints (\code{=:=})
to the results
where the Boolean equality operator (\code{==}) is used (which
does not perform bindings but enumerate all values).
As expected, the creation and traversal of a large search space introduced
by \code{==} is in most cases much slower than our presented approach
with variable bindings.
In addition, the example \code{last} shows that the lazy 
unification operator (\code{=:<=}) improves performance when unifying an
expression which has to be evaluted only partially. Using strict unification,
all elements of the list are (unnecessarily) evaluated. Interestingly, in the
\code{varInExp} example equational constraints perform significantly worse
than the other equality operators. \todo{Explain why}

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & KiCS2 & PAKCS & MCC  \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.52  & 4.27   & 0.15 \\
% \code{grep}                             & 0.13  & 2.83   & 0.15 \\
\code{simplify}                         & 10.33 & 0.15   & 0.01 \\
\code{varInExp}                         & 86.27 & 0.35   & 0.03 \\
\code{fromPeano (half (toPeano 10000))} & 6.26  & 109.71 & 3.17 \\
\code{palindrome}                       & 13.45 & 32.82  & 1.04 \\
% \code{horseman}                         & 4.93  & 8.85   & 0.42 \\
\hline
\end{tabular}
\caption{Benchmarks: equational constraints in different Curry implementations}
 \label{fig:unification}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|}
\hline
Expression                              & KiCS2 & PAKCS    \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.20  & 3.93     \\
\code{simplify}                         & 7.09  & 0.26     \\
\code{varInExp}                         & 0.14  & 0.69     \\
\code{fromPeano (half (toPeano 10000))} & 11.57 & $\infty$ \\
\code{palindrome}                       & 17.55 & $\infty$ \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different implementations of functional patterns}
 \label{fig:functional-patterns}
\end{figure}

As MCC does not support functional patterns, the performance of lazy
unification is compared with PAKCS only (Fig.~\ref{fig:functional-patterns}).
Again, our compiler performs well against PAKCS
and outperforms it in most cases (``$\infty$'' denotes a run time
of more than 30 minutes). \todo{extend}

>>>>

In general, to extract all values from a \code{Choice} structure,
we have to traverse it, compute all possible choices
but consider the choice identifiers to make consistent (left/right)
decisions.
The latter requirement can be implemented by storing
the decisions already made for some choices during the traversal.
For this purpose, we introduce the type
\begin{haskell}
  data Decision = NoDecision | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoDecision} represents the fact that the value of a choice
is yet undecided.
Furthermore, we assume operations to lookup the current decision
for a given identifier or change its decision (depending on the implementation
of \code{IDSupply}, KiCS2 supports various implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupDecision :: ID -> IO Decision
  setDecision    :: ID -> Decision -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following I/O operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
   where
    follow ChooseLeft  = printValsDFS x1
    follow ChooseRight = printValsDFS x2
    follow NoDecision  = do newDecision ChooseLeft  x1
                            newDecision ChooseRight x2

    newDecision d x = do setDecision i d
                         printValsDFS x
                         setDecision i NoDecision
  printValsDFS v                = print v
\end{haskell}
This operation ignores failures and prints values that are not a \code{Choice}.
For a \code{Choice}, it checks whether a decision for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoDecision}).
If a decision has been made for this choice, it follows this decision.
Otherwise, the left alternative is used and this decision is stored.
After printing
all the values w.r.t.\ this decision,
the decision is undone (like in backtracking)
and the right alternative is used and stored.

@@

  Fail         & _     =  Fail
  Choice i l r & c     =  Choice i (l & c) (r & c)
  _            & _     =  Fail

@@

% The benchmark results are from unif@climens2011_7_5_20_38_47.bench

In this section we evaluate our implementation by some benchmarks.
They were executed on a Linux machine running Debian 5.0.7 with
an Intel Core 2 Duo (3.0GHz) processor.
KiCS2 has been executed with
the Glasgow Haskell Compiler (GHC 7.0.4, option -O2) and uses
an efficient \code{IDSupply} implementation that makes use of \code{IORef}s.
For a comparison with other mature implementations of Curry,
we considered PAKCS \cite{Hanus10PAKCS}
(Version 1.9.2, based on a SICStus-Prolog 4.1.2)
and MCC \cite{Lux99FLOPS} (Version 0.9.10). The timings were performed with the
time command measuring the execution time (in seconds) of a compiled
executable for each benchmark as a mean of three runs.
The programs used for the benchmarks,
partially taken from \cite{AntoyHanus05LOPSTR},
are \code{last} (compute the last element of a list),\footnote{%
\ccode{inc x n} is a naive addition that \code{n} times
increases its argument \code{x} by 1.}
\code{simplify} (simplify a symbolic arithmetic expression),
\code{varInExp} (non-deterministically return a variable occuring
                  in a symbolic arithmetic expression),
\code{half} (compute the half of a peano number using logic variables),
and
\code{palindrome} (check whether a list is a palindrome).

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & \code{==~}& \code{=:=} & \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..100000])}   & 134.23    & 0.83      & 0.26        \\
% \code{grep}                             & 1.06      & 0.19      &             \\
\code{simplify}                         & 8.50      & 7.15      & 7.34        \\
\code{varInExp}                         & 11.72     & 1.50      & 1.40        \\
% \code{varInExp2}                        & 423.01    & 10.50     & 1.04        \\
% \code{varInExp3}                        & 399.68    & 16.70     & 11.28       \\
\code{fromPeano (half (toPeano 10000))} & 32.92     & 6.35      & 11.88       \\
\code{palindrome}                       & 28.59     & 15.58     & 23.21       \\
% \code{horseman}                         & 3.58      & 3.13      &             \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different equation solving methods}
 \label{fig:search-strategies}
\end{figure}

In Section~\ref{sec:Unification} we mentioned that equational
constraints could also be solved by generators without variable bindings,
but this technique might increase the search space due to the
possibly superfluous generation of all values.
To show the beneficial effects of our implementation
of equational constraints with variable bindings,
we compare in Fig.~\ref{fig:search-strategies}
the results of using equational constraints (\code{=:=})
to the results
where the Boolean equality operator (\code{==}) is used (which
does not perform bindings but enumerate all values).
As expected, the creation and traversal of a large search space introduced
by \code{==} is in most cases much slower than our presented approach
with variable bindings.
In addition, the example \code{last} shows that the lazy 
unification operator (\code{=:<=}) improves performance when unifying an
expression which has to be evaluted only partially. Using strict unification,
all elements of the list are (unnecessarily) evaluated.

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & KiCS2 & PAKCS & MCC  \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.83  & 4.25   & 0.14 \\
% \code{grep}                             & 0.19  & 2.88   & 0.14 \\
\code{simplify}                         & 7.15  & 0.15   & 0.00 \\
\code{varInExp}                         & 1.50  & 3.89   & 0.30 \\
% \code{varInExp2}                        & 10.50 & 3.48   & stckOflw \\
% \code{varInExp3}                        & 16.70 & 19.22  & 1.18 \\
\code{fromPeano (half (toPeano 10000))} & 6.35  & 108.88 & 3.22 \\
\code{palindrome}                       & 15.58 & 32.56  & 1.07 \\
% \code{horseman}                         & 3.13  & 8.70   & 0.42 \\
\hline
\end{tabular}
\caption{Benchmarks: equational constraints in different Curry implementations}
 \label{fig:unification}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|}
\hline
Expression                              & KiCS2 & PAKCS    \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.26  & 3.97     \\
\code{simplify}                         & 7.34  & 0.27     \\
\code{varInExp}                         & 1.40  & 8.88     \\
% \code{varInExp2}                         & 1.04  & 0.08    \\
% \code{varInExp3}                         & 11.28 & 16.62   \\
\code{fromPeano (half (toPeano 10000))} & 11.88 & $\infty$ \\
\code{palindrome}                       & 23.21 & $\infty$ \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different implementations of functional patterns}
 \label{fig:functional-patterns}
\end{figure}

As MCC does not support functional patterns, the performance of lazy
unification is compared with PAKCS only (Fig.~\ref{fig:functional-patterns}).
Again, our compiler performs well against PAKCS
and outperforms it in most cases (``$\infty$'' denotes a run time
of more than 30 minutes).
